---
title: "Penguin assignment MT22"
output: html_document
Purpose of script: 
##      Loads and clean penguin data for
##      linear regresion and piecewise regression analysis
##      of gentoo penguin bill length and body mass
---

# Setting up

Set the working directory:
```{r}
setwd("/my/file/path")
```

Sourcing the code for packages, cleaning the code and plotting:
```{r, message = FALSE}
source("functions/libraries.r")
source("functions/cleaningthecode.r")
source("functions/plotting.r")
```

# Loading and cleaning the data for analysis

Loading the raw data:
```{r}
penguins_raw <- read.csv("data_raw/penguins_raw.csv")
```

Using the cleaning function to remove capital letters and unwanted columns:
```{r}
penguins_clean <- cleaning(penguins_raw)
```

Looking at the column titles to see if the right ones have been removed and confirm lower case:
```{r}
names(penguins_clean)
```

Subsetting the cleaned data set for the columns necessary for the analysis and just the gentoo rows:
```{r}
gentoo_massbill <- bodymass_bills(penguins_clean)
```

Looking at the first few rows to confirm that the subsetting function has worked:
```{r}
head(gentoo_massbill)
```
Saving this code as data for analysis:
```{r}
write.csv(gentoo_massbill, paste0("data_for_analysis/gentoo_massbill.csv"))
```

***

# Linear regression analysis of gentoo penguins' bill length and body mass

We are going to see how well one numerical variable can predict the other. 

    Null hypothesis - the slope of the regression will equal zero
    Alternative hypothesis - the slope of the regression will not equal zero
    

Making a linear model with gentoo bill (AKA culmen) length as the explanatory variable and body mass as the response variable:
```{r}
mod_gentoo <- lm(body_mass_g ~ culmen_length_mm, gentoo_massbill)
```

Looking at the summary of the model:
```{r}
summary(mod_gentoo)
```

**Discussion:**

Looking at the coefficient table, the estimates can be used to write a linear equation where x = culmen length and y = body mass:

$y = 109.46x - 123.83$

* The estimate in the first row is the y intercept (= -123.83)
    + "Biological" interpretation: if a Gentoo had no bill, they would weigh -123.83g 
* The estimate in the second row is the slope of the line (= 109.46)
    + "Biological" interpretation: for every mm growth in the bill, there's a 109.46 g increase in the         body mass of the gentoo
    + The slope is significantly different from 0 (p value is less than 0.001)
    
The "Adjusted R-squared" value is also biologically relevant:
44.3% of Gentoo body mass is explained by culmen length, which is relatively high for biological research


## Plotting the linear regression


```{r, results='hide', message = FALSE}
gentoo_plot <- plot_gentoo_figure(gentoo_massbill)
gentoo_plot
```

**Discussion**

The linear regression line is the blue line and the black dots are the actual data points. Up until around a 50 mm bill length, the black dots lie on both sides of the linear regression. After 50 mm bill length, body mass seems to plateau and the area within the 95% confidence interval expands. The next section of this script will consider how well a piecewise regression can fit the data. 

***

# Piecewise regression analysis of gentoo penguins' bill length and body mass

Making the the piecewise regression model using segmented() from tidyverse:
```{r}
mod_segmented <- segmented(mod_gentoo)
summary(mod_segmented)
```
**Discussion**

The estimated break point in the piecewise (2-segment) regression is at around 50 mm bill length, with a standard error of 1.6mm either side, which matches the observations in the previous discussion. 

Looking at the coefficient table, the slopes of the two segments can be compared:
* The estimate in the first row represents the Y intercept of the first segment ( = -1136.4)
* The estimate in the second row represents the slope of the first segment (= 131.39)
* The estimate in the third row represents the difference in slope between the first and second segment of the regression (= -79.38; slope = 131.39 - 79.38 = 52.01)
    + Biological interpretation: at 50 mm bill length, Gentoo penguin body mass increases less per mm        increase in bill length

The "Adjusted R-squared" value is also biologically relevant:
45.5% of Gentoo body mass is explained by culmen length. This is slightly higher value than the first linear regression.

## Plotting the piecewise regression

Making a new data frame with an added column that has the fitted values and saving it to the data for analysis folder:
```{r}
gentoo_segmented_data <- add_column(gentoo_massbill, fitted_seg1 = mod_segmented$fitted.values)
write.csv(gentoo_segmented_data, paste0("data_for_analysis/gentoo_segmented_data.csv"))
```

Plotting the piecewise regression:
```{r}
segmented_plot <- plot_segmented_figure(gentoo_segmented_data)
segmented_plot
```
**Discussion**

Now, at bill lengths higher than 50 mm, data points lie on both sides of the regression line. Further analysis can be done using qqplots to compare the residuals to the fitted line (not included in this script).

# Brief conclusions

While the piecewise function explains slightly more variation than the linear regression, there still
need to be biological support. For example, can a developmental change be linked to a 50 mm bill? Also, there were fewer points at higher bill lengths, which will influence how the line is fitted. Further sampling of gentoo penguin populations can provide more data points to consider in fitting a model.

# Saving outputs

Saving scatter plot with regression line with sizing appropriate for a poster:
```{r, results='hide', message = FALSE}
save_gentoo_plot_png(gentoo_massbill, "out/figures/linear_plot.png", 
                      size = 17, res = 600, scaling = 1)
```


Saving the scatter plot with the piecewise regression with sizing appropriate for a poster:
```{r, results='hide', message = FALSE}
save_segmented_plot_png(gentoo_segmented_data, "out/figures/piecewise_plot.png", 
                      size = 17, res = 600, scaling = 1)
```

Saving the linear regression summary:
```{r, results='hide', message = FALSE}
sink(file = "out/linear-regression-summary.txt")
summary(mod_gentoo)
sink()
```

Saving the piecewise regression summary:
```{r, results='hide', message = FALSE}
sink(file = "out/pw-regression-summary.txt")
summary(mod_segmented)
sink()
```

Saving the software versions used:
```{r, results='hide', message = FALSE}
sink(file = "out/package-versions.txt")
sessionInfo()
sink()
```
